Fail Faster

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md

Reviewer A, Q1:
The main novelty of this work lies in applying staging to PBT. While staging
techniques are well established, their application to the abstractions and
performance bottlenecks of PBT libraries has not been explored, nor is it
straightforward. Our paper is intended to promote staging as a tool in the PBT
developer’s toolkit by showing that it can erase abstraction overhead, leading
to substantial performance improvements.

Prior to this work, neither abstraction overhead nor
sampling costs were viewed as major targets for optimization in PBT. As
evidence, we point to SOTA libraries like `Base_quickcheck`, which is highly
performance sensitive yet retains design choices (monadic combinators, expensive
randomness) that introduce significant overhead.

Additional novelty lies in Section 3.7, where we stage type-derived generators.
Unlike most staged libraries, which require users to understand metaprogramming,
our approach is fully automatic. Since type-derived generators are constructed
at compile time, they can be staged without altering user experience. This is a
rare example of staging "for free."

---

Reviewer B also asks (Q2) if the testing-time speedups are a benefit, given (a) the time required to compile the staged code, and (b) the relative balance between input-generating and property-running durations in practice.

These are both excellent points...

For (a), we do not believe that this is an issue in practice, for two reasons.

1. Generator code changes extremely infrequently compared to code being
tested, and the generator metaprogram needs to be re-run and the output code
recompiled only as the generator itself changes. This is somewhat different from
many other uses of run-time metaprogramming, where the application code itself is
written as a metaprogram. Our intended mode of usage therefore is more like compile time metaprogramming
than run-time metaprogramming, where programmers save the code output to disk to cache it.
2. Generators are quite small, and compile quite quickly.

<!-- ``The authors need to use more workloads and justify this more clearly.'' You're saying... give us *more* workloads? -->

<!-- These benchmarks are relatively challenging to build: one needs many a program, many properties, and many mutants of varying known difficulties -->

<!-- 
The second criticism is with the evaluation. It is unclear to me if, in the software development cycle, the proposed improvement will provide any actual benefit for two main reasons. First, the compiler time is not taken into account. Multi-stage programming adds a runtime code generation time; the extent to which it is taken into account is unclear. It has an overhead that can be amortized if the computation itself is very time-consuming. The authors need to clarify if this time is taken into account. Second, based on Amdahl's Law, does the proposed improvement bring actual benefits in practice? RQ2 shows results for finding bugs in the Etna platform. However, it is still unclear how many practical applications one would see in which a massive proportion of overhead for the testing infrastructure vs. the time each function itself takes. The authors need to use more workloads and justify this more clearly. -->

---

Reviewer C:

1. Can you provide an example of how the recursive generator API in Section 3.6 is used?
Yes, it's used in Figure 3 --- we can include a figure down at section 3.6 with a small example in the camera ready.

2. How were the experimental benchmarks chosen?
We used the subset of the pre-existing Etna suite of benchmarks. Etna was
designed exactly for this purpose: to test the bugfinding capabilities of
competing PBT generators.  There are two benchmarks from the original Etna we
did not use. One tests Red-Black trees, and the other tests a lambda calculus
typed with System FSub, a more sophisticated type system.  We felt that the RBT eval was
redundant with the BTS eval we already used: the generators are very similar.
We chose not to use the FSub eval for two reasons. First, it had not been implemented
in OCaml, and second, we felt this would be redundant with the STLC
evals. The FSub generator is even more complex with more combinators than the STLC one, but the testing code and properties are essentially the same:
we would see an exaggerated version of the speedups in the STLC eval.

3. What do you mean by "semantically identical", and are the generators also syntactically similar?

Yes, the AllegrOCaml and Base_quickcheck generors are as you imagine: essentially syntactically identical, modulo
some extra syntactic cruft for staging. We plan to submit all of our code with the artifact evaluation, but for now, we
include an example from our evals of a generator and its staged counterpart below the fold.

By "semantically identical", we mean that given the same random seeds, the two
generators produce the same generated output.  This property is important
because it guarantees that all of our bug-finding speedups are solely
attributable to the improved performance of staged generators, and not because
the staged generators happen to luckily create bug-squashing outputs. The
sequence of outputs is the same between the staged and unstaged versions, the staged
generators just produce them faster.

We further note that this is not an obvious or simple property to maintain:
staged and unstaged programs can look syntactically similar but behave
non-equivalently, especially in the presence of effects (for example, `let x =
print "hi" in x;x` and `let x = .<print "hi">. in .<$x;$x>. --- the former
prints once, the latter twice). Section 3.4 is essentially about ensuring that
programmers need not worry about this, and can write their generators in a
manner that is syntactically as similar as possible to the non-staged versions.

```
let rec gen ~(lo: int) ~(hi: int)  =
  let%bind sz = size in
  let should_stop = lo >= hi || sz <= 1  in
  if should_stop
    then return E
  else
    weighted_union [
      (1., return E);
      (float_of_int sz , (
        let%bind k = int_inclusive ~lo ~hi in
        let%bind v = Nat.quickcheck_generator_parameterized bst_bespoke_limits in
        let%bind left = with_size ~size_c:(sz / 2) (gen ~lo:lo ~hi:(k - 1))  in
        let%bind right = with_size ~size_c:(sz / 2) (gen ~lo:(k + 1) ~hi:hi) in
        return (T (left, k, v, right))  
      ))
    ]

let staged_quickcheck_generator (lo: int code) (hi: int code) : tree code G.t =
  recursive (.< (.~lo, .~hi ) >.) 
  (fun go lohi -> 
    let%bind sz = size in
    let%bind (lo, hi) = split_pair lohi in
    let%bind should_stop = split_bool .< .~hi <= .~lo || .~sz <= 1 >. in
    if should_stop
      then return .< E >.
    else
      weighted_union [
        (.< 1. >., return .< E >.);
        ((G.C.i2f sz), (
          let%bind k = int_inclusive ~lo ~hi in
          let%bind v = Nat.staged_quickcheck_generator_sr_t (G.C.lift bst_bespoke_limits) in
          let%bind left = with_size ~size_c:(G.C.div2 sz) (recurse go .<(.~lo, .~k - 1) >.) in
          let%bind right = with_size ~size_c:(G.C.div2 sz) (recurse go .<(.~k + 1, .~hi) >.) in
          return (.< T (.~left, .~k, .~v, .~right) >.)))
      ]
  )
```

-------------

Reviewer B:

> Typos on l379, l475, l621, l653, l718, l824

p4: It's odd that Figure 4 appears before Figure 3.

: "returns a code" should be "returns code"

: "a observation"

: "a 'a code" - perhaps you mean "a value of type 'a code"?

> l119: I'm curious whether OCaml 5 might benefit from using it's native
effect handlers in place of a monadic generator DSL.

We are also now curious about this! Very interesting idea, direct-style generators
could potentially be very performant and also have the benefit of being more idiomatic.

> l989: You have conjectured that because GHC is set up to perform aggressive optimisations without staging then it is likely to benefit less from the kind of optimisations you exploit. It would be worth investigating to what extent this is indeed the case. Given that GHC’s QuickCheck is the canonical PBT framework it seems particularly worthwhile to perform further experiments with it. I wonder to what extent it would be possible to disable some of GHC’s aggressive inlining, both in order to assess how much it is really paying off, and to compare its robustness to your staging approach.

This is a fair point. Our conjecture is based on anecdotal evidence; one of the authors — an experienced Haskell developer — manually applied some of Allegro’s optimizations to a few Haskell generators, and they found that they could really only manage to make performance worse, not better. This is nowhere near a proof, but it discouraged us from exploring that path in the short term. We would be happy to mention this anecdotal experience and/or go into more detail about wanting to do this experiment as future work

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md

## Reviewer A

### Q1) Can you clarify the novelty of your paper? Does it introduce any optimization from the multi-stage programming not known before?

The main novelty of this work lies in applying staging to PBT. While staging
techniques are well established, their application to the abstractions and
performance bottlenecks of PBT libraries has not been explored, nor is it
straightforward. Our paper is intended to promote staging as a tool in the PBT
developer’s toolkit by showing that it can erase abstraction overhead, leading
to substantial performance improvements.

@@

Prior to this work, neither abstraction overhead nor
sampling costs were viewed as major targets for optimization in PBT. As
evidence, we point to SOTA libraries like `Base_quickcheck`, which is highly
performance sensitive yet retains design choices (monadic combinators, expensive
randomness) that introduce significant overhead.

Additional novelty lies in Section 3.7, where we stage type-derived generators.
Unlike most staged libraries, which require users to understand metaprogramming,
our approach is fully automatic. Since type-derived generators are constructed
at compile time, they can be staged without altering user experience. This is a
rare example of staging "for free."

### Q2) Can you elaborate on the end-to-end effectiveness of your proposed solution?

The end-to-end effectiveness of our solution depends on (a) the complexity 
of the inputs being generated, and (b) the amount of time generation takes 
relative to other parts of the testing process—most obviously, running the 
system on the generated input.

In general, more complex generators contain more calls to bind that are 
fused away using the Allegro technique; therefore, more complex generators 
see a larger speedup. Figure 15 showcases this relationship.

The effectiveness of efficient input generation on total testing time is 
dependent on how long everything that isn’t input generation takes. In systems 
where running the tests themselves is slow, speeding up input generation won’t 
have as large an impact; in systems where testing is fast, it will constitute a 
major speedup. Our end-to-end tests on bug-finding speed, shown in Fig. 17 and 
Fig. 18, show the time it takes for PBTs to find bugs in mutated programs. In 
these examples, faster input generation has a measurable end-to-end impact on 
all of our benchmarks—up to 2.65X using staging alone, and 3.40X in combination 
with fast randomness.

### Q3) What would it take to implement type-derived generators in Scala? Why did you decide not to implement it?

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md

We are also now curious about this! Very interesting idea, direct-style generators
could potentially be very performant and also have the benefit of being more idiomatic.

> l989: You have conjectured that because GHC is set up to perform aggressive optimisations without staging then it is likely to benefit less from the kind of optimisations you exploit. It would be worth investigating to what extent this is indeed the case. Given that GHC’s QuickCheck is the canonical PBT framework it seems particularly worthwhile to perform further experiments with it. I wonder to what extent it would be possible to disable some of GHC’s aggressive inlining, both in order to assess how much it is really paying off, and to compare its robustness to your staging approach.

This is a fair point. Our conjecture is based on anecdotal evidence; one of the authors — an experienced Haskell developer — manually applied some of Allegro’s optimizations to a few Haskell generators, and they found that they could really only manage to make performance worse, not better. This is nowhere near a proof, but it discouraged us from exploring that path in the short term. We would be happy to mention this anecdotal experience and/or go into more detail about wanting to do this experiment as future work

# Reviewer C

> 1. Can you provide an example of how the recursive generator API in Section 3.6 is used?
Yes, it's used in Figure 3 --- we can include a figure down at section 3.6 with a small example in the camera ready.

> 2. How were the experimental benchmarks chosen?

We used the subset of the pre-existing Etna suite of benchmarks. Etna was
designed exactly for this purpose: to test the bugfinding capabilities of
competing PBT generators.  There are two benchmarks from the original Etna we
did not use. One tests Red-Black trees, and the other tests a lambda calculus
typed with System FSub, a more sophisticated type system.  We felt that the RBT eval was
redundant with the BTS eval we already used: the generators are very similar.
We chose not to use the FSub eval for two reasons. First, it had not been implemented
in OCaml, and second, we felt this would be redundant with the STLC
evals. The FSub generator is even more complex with more combinators than the STLC one, but the testing code and properties are essentially the same:
we would see an exaggerated version of the speedups in the STLC eval.

> 3. What do you mean by "semantically identical", and are the generators also syntactically similar?

> 4. Why was BST (Repeated Insert) not used for the ScAllegro experiments in Figure 16?

We picked a representative from each category of generators: trees, lists, and STLC terms, representing low, medium, and high numbers of binds, respectively. We anticipate BST (Repeated Insert) would see a slightly smaller speedup due to its lower number of binds, and we would be happy to include this information in a revision.

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md

The main novelty of this work lies in applying staging to PBT. While staging
techniques are well established, their application to the abstractions and
performance bottlenecks of PBT libraries has not been explored, nor is it
straightforward. Our paper is intended to promote staging as a tool in the PBT
developer’s toolkit by showing that it can erase abstraction overhead, leading
to substantial performance improvements.

Prior to this work, neither abstraction overhead nor
sampling costs were viewed as major targets for optimization in PBT. As
evidence, we point to SOTA libraries like `Base_quickcheck`, which is highly
performance sensitive yet retains design choices (monadic combinators, expensive
randomness) that introduce significant overhead.

Additional novelty lies in Section 3.7, where we stage type-derived generators.
Unlike most staged libraries, which require users to understand metaprogramming,
our approach is fully automatic. Since type-derived generators are constructed
at compile time, they can be staged without altering user experience. This is a
rare example of staging "for free."

> Q2) Can you elaborate on the end-to-end effectiveness of your proposed solution?

@@

The end-to-end effectiveness of our solution depends on (a) the complexity 
of the inputs being generated, and (b) the amount of time generation takes 
relative to other parts of the testing process—most obviously, running the 
system on the generated input.

In general, more complex generators contain more calls to bind that are 
fused away using the Allegro technique; therefore, more complex generators 
see a larger speedup. Figure 15 showcases this relationship.

The effectiveness of efficient input generation on total testing time is 
dependent on how long everything that isn’t input generation takes. In systems 
where running the tests themselves is slow, speeding up input generation won’t 
have as large an impact; in systems where testing is fast, it will constitute a 
major speedup. Our end-to-end tests on bug-finding speed, shown in Fig. 17 and 
Fig. 18, show the time it takes for PBTs to find bugs in mutated programs. In 
these examples, faster input generation has a measurable end-to-end impact on 
all of our benchmarks—up to 2.65X using staging alone, and 3.40X in combination 
with fast randomness.

> Q3) What would it take to implement type-derived generators in Scala? Why did you decide not to implement it?

@@

<!-- ``The authors need to use more workloads and justify this more clearly.'' You're saying... give us *more* workloads? -->

<!-- These benchmarks are relatively challenging to build: one needs many a program, many properties, and many mutants of varying known difficulties -->

<!-- 
The second criticism is with the evaluation. It is unclear to me if, in the software development cycle, the proposed improvement will provide any actual benefit for two main reasons. First, the compiler time is not taken into account. Multi-stage programming adds a runtime code generation time; the extent to which it is taken into account is unclear. It has an overhead that can be amortized if the computation itself is very time-consuming. The authors need to clarify if this time is taken into account. Second, based on Amdahl's Law, does the proposed improvement bring actual benefits in practice? RQ2 shows results for finding bugs in the Etna platform. However, it is still unclear how many practical applications one would see in which a massive proportion of overhead for the testing infrastructure vs. the time each function itself takes. The authors need to use more workloads and justify this more clearly. -->

> One somewhat open-ended question I am wondering about is to what
extent we can bridge the performance gap between bespoke
highly-optimised fuzzing approaches and the much more generic PBT
libraries exemplified by QuickCheck. Can we hope to build fuzzers on
top of generic PBT libraries which perform as well as the hand-coded
fuzzers used in practice? Does your work shed any light on how to
tackle this question?

This is a really interesting question, and we think that our work speaks directly to part of it. Specifically, hand-coded fuzzers that construct inputs satisfying specific structural and semantic constraints are performing precisely the same task that PBT generators are designed for. Hand-coded fuzzers might be faster than PBT generators if users choose to optimize them aggressively, but ideally those optimizations should not need to be done manually. Our hope is that Allegro (and other advances from the PBT literature) can make PBT generators fast enough that developers could use them as highly-performant fuzzers, even though they are expressed in a higher-level language that is easier to write in and reason about. The question gets a bit less clear when considering fuzzers that use mutation and other techniques to obtain interesting inputs, rather than a hand-coded program, although ideas in papers like Coverage Guided Property-Based Testing and Parsing Randomness may provide paths forward in those domains as well.

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md

Reviewer A:

> Q1) Can you clarify the novelty of your paper? Does it introduce any 
optimization from the multi-stage programming not known before?

The main novelty of this work lies in applying staging to PBT. While staging
techniques are well established, their application to the abstractions and
performance bottlenecks of PBT libraries has not been explored, nor is it
straightforward. Our paper is intended to promote staging as a tool in the PBT
developer’s toolkit by showing that it can erase abstraction overhead, leading
to substantial performance improvements.

Prior to this work, neither DSL abstraction overhead nor sampling costs were
viewed as targets for optimization in PBT. As evidence, we point to SOTA
libraries like `Base_quickcheck`, which is has had a lot of optimization effort
put into it --- it uses imperative features like mutable arrays and a stateful
RNG for performance reasons --- yet retains design choices (monadic combinators,
expensive randomness) that introduce significant overhead.

@@

Additional novelty lies in Section 3.7, where we stage type-derived generators.
Unlike most staged libraries, which require users to understand metaprogramming,
our approach is fully automatic. Since type-derived generators are constructed
at compile time, they can be staged without altering user experience. This is a
rare example of staging "for free."

>  Q2) Can you elaborate on the end-to-end effectiveness of your proposed 
solution?

[JWC: reframe this to be more about the eval.]

The effectiveness depends on (a) the complexity 
of the inputs being generated, and (b) the amount of time generation takes 
relative to other parts of the testing process—most obviously, running the 
system on the generated input.

@@

In general, more complex generators contain more calls to bind that are 
fused away using the Allegro technique; therefore, more complex generators 
see a larger speedup. Figure 15 showcases this relationship.

The effectiveness of efficient input generation on total testing time is 
dependent on how long everything that isn’t input generation takes. In systems 
where running the tests themselves is slow, speeding up input generation won’t 
have as large an impact; in systems where testing is fast, it will constitute a 
major speedup. Our end-to-end tests on bug-finding speed, shown in Fig. 17 and 
Fig. 18, show the time it takes for PBTs to find bugs in mutated programs. In 
these examples, faster input generation has a measurable end-to-end impact on 
all of our benchmarks—up to 2.65X using staging alone, and 3.40X in combination 
with fast randomness.

As for the compilation time of the emitted generator code, we do not believe
that this is an issue, for two reasons. The first is that our metaprogrammed
generators compile extremely quickly in AllegrOCaml: 33ms for the simple Bool List
generator, up to 54ms for the STLC generator (a full table of the AllegrOCaml generators and their compilation times can be found after the fold).  Amortized across testing multiple
properties (which frequently take on the order of seconds to run), this is
negligible. Second, generator code changes extremely infrequently compared to
code being tested, and the generator metaprogram needs to be re-run and the
output code recompiled only when the generator itself changes. This is somewhat
different from many other uses of run-time metaprogramming, where the
application code itself is written as a metaprogram. Our intended mode of usage
therefore is more like compile time metaprogramming than run-time
metaprogramming, where programmers save the code output to disk to cache it.

@@

: "a 'a code" - perhaps you mean "a value of type 'a code"?

> l119: I'm curious whether OCaml 5 might benefit from using it's native
effect handlers in place of a monadic generator DSL.

We are also now curious about this! Very interesting idea, direct-style generators
could potentially be very performant and also have the benefit of being more idiomatic.

> l989: You have conjectured that because GHC is set up to perform aggressive optimisations without staging then it is likely to benefit less from the kind of optimisations you exploit. It would be worth investigating to what extent this is indeed the case.

This is a fair point. Our conjecture is based on anecdotal evidence; one of the authors — an experienced Haskell developer — manually applied some of Allegro’s optimizations to a few Haskell generators, and they found that they could really only manage to make performance worse, not better. This is nowhere near a proof, but it discouraged us from exploring that path in the short term. We would be happy to mention this anecdotal experience and/or go into more detail about wanting to do this experiment as future work.

> Can we can bridge the performance gap between bespoke
highly-optimised fuzzing approaches and the much more generic PBT
libraries exemplified by QuickCheck?

@@

This is a really interesting question, and we think that our work speaks 
directly to part of it. Specifically, hand-coded fuzzers that construct 
inputs satisfying specific structural and semantic constraints are performing 
precisely the same task that PBT generators are designed for. Hand-coded fuzzers 
might be faster than PBT generators if users choose to optimize them aggressively, 
but ideally those optimizations should not need to be done manually. Our hope is 
that Allegro (and other advances from the PBT literature) can make PBT generators 
fast enough that developers could use them as highly-performant fuzzers, even 
though they are expressed in a higher-level language that is easier to write 
in and reason about. The question gets a bit less clear when considering fuzzers 
that use mutation and other techniques to obtain interesting inputs, rather than 
a hand-coded program, although ideas in papers like Coverage Guided Property-Based 
Testing and Parsing Randomness may provide paths forward in those domains as well.

----

Reviewer C:

> 1. Can you provide an example of how the recursive generator API in Section 3.6 is used?
Yes, it's used in Figure 3 --- we can include a figure down at Section 3.6 with a small example in the camera-ready version.

> 2. How were the experimental benchmarks chosen?
We used the subset of the pre-existing Etna suite of benchmarks. Etna was
designed exactly for this purpose: to test the bugfinding capabilities of
competing PBT generators.  There are two benchmarks from the original Etna we
did not use. One tests Red-Black trees, and the other tests a lambda calculus
typed with System FSub, a more sophisticated type system.  We felt that the RBT eval was
redundant with the BTS eval we already used: the generators are very similar.
We chose not to use the FSub eval for two reasons. First, it had not been implemented
in OCaml, and second, we felt this would be redundant with the STLC
evals. The FSub generator is even more complex with more combinators than the STLC one, but the testing code and properties are essentially the same:
we would see an exaggerated version of the speedups in the STLC eval.

@@

```
let rec gen ~(lo: int) ~(hi: int)  =
  let%bind sz = size in
  let should_stop = lo >= hi || sz <= 1  in
  if should_stop
    then return E
  else
    weighted_union [
      (1., return E);
      (float_of_int sz , (
        let%bind k = int_inclusive ~lo ~hi in
        let%bind v = Nat.quickcheck_generator_parameterized bst_bespoke_limits in
        let%bind left = with_size ~size_c:(sz / 2) (gen ~lo:lo ~hi:(k - 1))  in
        let%bind right = with_size ~size_c:(sz / 2) (gen ~lo:(k + 1) ~hi:hi) in
        return (T (left, k, v, right))  
      ))
    ]

let staged_quickcheck_generator (lo: int code) (hi: int code) : tree code G.t =
  recursive (.< (.~lo, .~hi ) >.) 
  (fun go lohi -> 
    let%bind sz = size in
    let%bind (lo, hi) = split_pair lohi in
    let%bind should_stop = split_bool .< .~hi <= .~lo || .~sz <= 1 >. in
    if should_stop
      then return .< E >.
    else
      weighted_union [
        (.< 1. >., return .< E >.);
        ((G.C.i2f sz), (
          let%bind k = int_inclusive ~lo ~hi in
          let%bind v = Nat.staged_quickcheck_generator_sr_t (G.C.lift bst_bespoke_limits) in
          let%bind left = with_size ~size_c:(G.C.div2 sz) (recurse go .<(.~lo, .~k - 1) >.) in
          let%bind right = with_size ~size_c:(G.C.div2 sz) (recurse go .<(.~k + 1, .~hi) >.) in
          return (.< T (.~left, .~k, .~v, .~right) >.)))
      ]
  )
```

> 4. Why was BST (Repeated Insert) not used for the ScAllegro experiments in Figure 16?

We picked a representative from each category of generators: trees, lists, and STLC terms, 
representing low, medium, and high numbers of binds, respectively. We anticipate BST 
(Repeated Insert) would see a slightly smaller speedup due to its lower number of binds, 
and we would be happy to include this information in a revision. Our goal in this evaluation subsection was not to point-by-point reproduce the AllegrOCaml evaluation in ScAllegro, but
simply to demonstrate that the performance improvments of the Allegro technique are portable
across languges.

----

Generator Compilation Times

Bool List: 33.183ms
BST Repated Insert: 45.453ms
BST Type: 46.051ms
BST Single Pass: 50.147ms
STLC Type: 44.663ms
STLC Bespoke: 53.699ms

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md

: "a 'a code" - perhaps you mean "a value of type 'a code"?

> l119: I'm curious whether OCaml 5 might benefit from using it's native
effect handlers in place of a monadic generator DSL.

We are also now curious about this! Very interesting idea, direct-style generators
could potentially be very performant and also have the benefit of being more idiomatic.

> l989: You have conjectured that because GHC is set up to perform aggressive optimisations without staging then it is likely to benefit less from the kind of optimisations you exploit. It would be worth investigating to what extent this is indeed the case. Given that GHC’s QuickCheck is the canonical PBT framework it seems particularly worthwhile to perform further experiments with it. I wonder to what extent it would be possible to disable some of GHC’s aggressive inlining, both in order to assess how much it is really paying off, and to compare its robustness to your staging approach.

This is a fair point. Our conjecture is based on anecdotal evidence; one of the authors — an experienced Haskell developer — manually applied some of Allegro’s optimizations to a few Haskell generators, and they found that they could really only manage to make performance worse, not better. This is nowhere near a proof, but it discouraged us from exploring that path in the short term. We would be happy to mention this anecdotal experience and/or go into more detail about wanting to do this experiment as future work

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md

# Reviewer A

> Q1) Can you clarify the novelty of your paper? Does it introduce any optimization from the multi-stage programming not known before?

[JWC: ... and noticing that these sources of inefficiency exist in PBT]

@@

Prior to this work, neither DSL abstraction overhead nor sampling costs were
viewed as targets for optimization in PBT. As evidence, we point to SOTA
libraries like `Base_quickcheck`, which is has had a lot of optimization effort
put into it --- it uses imperative features like mutable arrays and a stateful
RNG for performance reasons --- yet retains design choices (monadic combinators,
expensive randomness) that introduce significant overhead.

Additional novelty lies in Section 3.7, where we stage type-derived generators.
Unlike most staged libraries, which require users to understand metaprogramming,
our approach is fully automatic. Since type-derived generators are constructed
at compile time, they can be staged without altering user experience. This is a
rare example of staging "for free."

>  Q2) Can you elaborate on the end-to-end effectiveness of your proposed 
solution?

[JWC: reframe this to be more about the eval.]
<!-- 
The effectiveness depends on (a) the complexity 
of the inputs being generated, and (b) the amount of time generation takes 
relative to other parts of the testing process—most obviously, running the 
system on the generated input.

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md

The main novelty of this work lies in applying staging to PBT. While staging
techniques are well established, their application to the abstractions and
performance bottlenecks of PBT libraries has not been explored, nor is it
straightforward. Our paper is intended to promote staging as a tool in the PBT
developer’s toolkit by showing that it can erase abstraction overhead, leading
to substantial performance improvements.

Prior to this work, neither DSL abstraction overhead nor sampling costs were
viewed as targets for optimization in PBT. As evidence, we point to SOTA
libraries like `Base_quickcheck`, which is has had a lot of optimization effort
put into it --- it uses imperative features like mutable arrays and a stateful
RNG for performance reasons --- yet retains design choices (monadic combinators,
expensive randomness) that introduce significant overhead.

Additional novelty lies in Section 3.7, where we stage type-derived generators.
Unlike most staged libraries, which require users to understand metaprogramming,
our approach is fully automatic. Since type-derived generators are constructed
at compile time, they can be staged without altering user experience. This is a
rare example of staging "for free."

[JWC can we also say that this is an instance of two-level metaprogramming? It's a metaprogram that generates metaprograms!]

@@

The effectiveness of efficient input generation on total testing time is 
dependent on how long everything that isn’t input generation takes. In systems 
where running the tests themselves is slow, speeding up input generation won’t 
have as large an impact; in systems where testing is fast, it will constitute a 
major speedup. Our end-to-end tests on bug-finding speed, shown in Fig. 17 and 
Fig. 18, show the time it takes for PBTs to find bugs in mutated programs. In 
these examples, faster input generation has a measurable end-to-end impact on 
all of our benchmarks—up to 2.65X using staging alone, and 3.40X in combination 
with fast randomness. -->

As for the compilation time of the emitted generator code, we do not believe
that this is an issue, for two reasons. The first is that our metaprogrammed
generators compile extremely quickly in AllegrOCaml: 33ms for the simple Bool List
generator, up to 54ms for the STLC generator (a full table of the AllegrOCaml generators and their compilation times can be found after the fold).  Amortized across testing multiple
properties (which frequently take on the order of seconds to run), this is
negligible. Second, generator code changes extremely infrequently compared to
code being tested, and the generator metaprogram needs to be re-run and the
output code recompiled only when the generator itself changes. This is somewhat
different from many other uses of run-time metaprogramming, where the
application code itself is written as a metaprogram. Our intended mode of usage
therefore is more like compile time metaprogramming than run-time
metaprogramming, where programmers save the code output to disk to cache it.

> Q3) What would it take to implement type-derived generators in Scala? 
Why did you decide not to implement it?

None of the authors are Scala experts. We are much more familiar with OCaml, so we chose to focus our implementation efforts there. That said, we see no fundamental barrier to implementing type-derived staged generators in Scala. Indeed, the scalacheck-derived project already uses Scala's runtime reflection to type-derive standard (unstaged) ScalaCheck generators. Adapting this approach to emit ScAllegro would require engineering effort, but we anticipate no fundamental obstacles beyond that.

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md

The effectiveness of efficient input generation on total testing time is 
dependent on how long everything that isn’t input generation takes. In systems 
where running the tests themselves is slow, speeding up input generation won’t 
have as large an impact; in systems where testing is fast, it will constitute a 
major speedup. Our end-to-end tests on bug-finding speed, shown in Fig. 17 and 
Fig. 18, show the time it takes for PBTs to find bugs in mutated programs. In 
these examples, faster input generation has a measurable end-to-end impact on 
all of our benchmarks—up to 2.65X using staging alone, and 3.40X in combination 
with fast randomness. -->

As for the compilation time of the emitted generator code, we do not believe
that this is an issue, for two reasons. The first is that our metaprogrammed
generators compile extremely quickly in AllegrOCaml: 33ms for the simple Bool List
generator, up to 54ms for the STLC generator (a full table of the AllegrOCaml generators and their compilation times can be found after the fold).  Amortized across testing multiple
properties (which frequently take on the order of seconds to run), this is
negligible. Second, generator code changes extremely infrequently compared to
code being tested, and the generator metaprogram needs to be re-run and the
output code recompiled only when the generator itself changes. This is somewhat
different from many other uses of run-time metaprogramming, where the
application code itself is written as a metaprogram. Our intended mode of usage
therefore is more like compile time metaprogramming than run-time
metaprogramming, where programmers save the code output to disk to cache it.

> Q3) What would it take to implement type-derived generators in Scala? 
Why did you decide not to implement it?

None of the authors are Scala experts: we are much more familiar with OCaml, so
we chose to focus our implementation efforts there. That said, we see no
fundamental barrier to implementing type-derived staged generators in Scala.
Indeed, the scalacheck-derived project already uses Scala's reflection features
to type-derive standard (unstaged) ScalaCheck generators. Adapting this approach
to emit ScAllegro would require engineering effort, but we anticipate no
fundamental obstacles beyond that.

@@

Thank you! We'll fix these for the camera ready.

> l119: I'm curious whether OCaml 5 might benefit from using it's native
effect handlers in place of a monadic generator DSL.

We are also now curious about this! Very interesting idea, direct-style generators
could potentially be very performant and also have the benefit of being more idiomatic.

> l639: Testing is all well and good, but have you not also considered
trying to construct a more rigorous proof that AllegrOCaml is correct
by program transformation / calculation?

We had not considered this, but it would be interesting to investigate.

> l751: Given that MetaOCaml is compatible with OCaml 5, why are you
still using OCaml 4?

We were not aware that MetaOCaml was now compatible with OCaml 5, this is good
to know!  When we began the project earlier this year, the MetaOCaml homepage
instructed users to install a version compatible with OCaml 4.14.1, which is
what we did (it has since been updated to recommend 5.3.0).

> l824: The text reads a bit oddly to me at this point, as you are
talking about an experiment you performed in the past and yet you
insist on using the present tense.

Thanks for the feedback, we'll workshop this prose to make it less awkward.

### 2025-12-02 paper/reviews/rebuttal_oopsla26.md

[JWC: reframe this to be more about the eval.]
<!-- 
The effectiveness depends on (a) the complexity 
of the inputs being generated, and (b) the amount of time generation takes 
relative to other parts of the testing process—most obviously, running the 
system on the generated input.

In general, more complex generators contain more calls to bind that are 
fused away using the Allegro technique; therefore, more complex generators 
see a larger speedup. Figure 15 showcases this relationship.

The effectiveness of efficient input generation on total testing time is 
dependent on how long everything that isn’t input generation takes. In systems 
where running the tests themselves is slow, speeding up input generation won’t 
have as large an impact; in systems where testing is fast, it will constitute a 
major speedup. Our end-to-end tests on bug-finding speed, shown in Fig. 17 and 
Fig. 18, show the time it takes for PBTs to find bugs in mutated programs. In 
these examples, faster input generation has a measurable end-to-end impact on 
all of our benchmarks—up to 2.65X using staging alone, and 3.40X in combination 
with fast randomness. -->

Regarding end-to-end effectiveness: Etna is the standard platform for 
evaluating the bug-finding speed of PBT generators. It was designed 
as a testbench for comparing generator strategies against each other, and, 
to our knowledge, no more comprehensive benchmark suite exists for this 
purpose. Our results show consistent speedups across Etna's tasks, with 
generation time translating directly to faster bug-finding.

We acknowledge that characterizing the proportion of testing time spent 
on generation across real-world PBT workloads is an interesting question 
that merits further study, and doing so would be an important research 
contribution in itself; we view this as future work outside 
the scope of this paper.
